Live coding

### 2019-01-17 wip-live-programming/paper.tex
Updating formalisation, formatting tweaks

\usepackage{booktabs}
\usepackage{subcaption}
\usepackage{lineno,hyperref,xcolor}
\usepackage{flushend}
\usepackage{stmaryrd}
\usepackage{amssymb}
\usepackage{xypic}
\usepackage{semantic}
\usepackage{booktabs} 
\usepackage{subcaption}
\usepackage{enumitem}
\usepackage{enumerate}
\usepackage{amsmath}

@@

\definecolor{cmtclr}{rgb}{0.0,0.6,0.0}
\definecolor{kvdclr}{rgb}{0.0,0.0,0.6}
\definecolor{numclr}{rgb}{0.0,0.4,0.0}
\definecolor{strclr}{rgb}{0.4,0.4,0.0}
\definecolor{rstrclr}{rgb}{0.5,0.1,0.0}
\definecolor{prepclr}{rgb}{0.6,0.0,0.2}
\newcommand{\vect}[1]{\langl #1 \rangl}
\newcommand{\langl}{\begin{picture}(4.5,7)
\put(1.1,2.5){\rotatebox{60}{\line(1,0){5.5}}}
\put(1.1,2.5){\rotatebox{300}{\line(1,0){5.5}}}
\end{picture}}
\newcommand{\rangl}{\begin{picture}(4.5,7)
\put(.9,2.5){\rotatebox{120}{\line(1,0){5.5}}}
\put(.9,2.5){\rotatebox{240}{\line(1,0){5.5}}}
\end{picture}}
\newcommand{\ball}[1]{\FPeval{\result}{clip(201+#1)}\textnormal{\ding{\result}}}
\newcommand{\lsep}{\;\;|\;\;}
\newcommand{\num}[1]{\textcolor{numclr}{#1}}
\newcommand{\str}[1]{\textnormal{\textcolor{strclr}{\sffamily "#1"}}}
\newcommand{\rstr}[1]{\textnormal{\textcolor{rstrclr}{\sffamily "#1"}}}
\newcommand{\ident}[1]{\textnormal{\sffamily #1}}
\newcommand{\qident}[1]{\textnormal{\sffamily \textquotesingle #1\textquotesingle}}
\newcommand{\dom}{\ident{dom}}
\newcommand{\kvd}[1]{\textnormal{\textcolor{kvdclr}{\sffamily #1}}}

\newcommand{\bndclr}[1]{\textcolor{kvdclr}{#1}}
\newcommand{\bkndclr}[1]{\textcolor{prepclr}{#1}}
\newcommand{\blblclr}[1]{\textcolor{numclr}{#1}}
\newcommand{\bnd}[1]{\textnormal{\textcolor{kvdclr}{\sffamily #1}}}
\newcommand{\bknd}[1]{\textnormal{\textcolor{prepclr}{\sffamily #1}}}
\newcommand{\blbl}[1]{\textnormal{\textcolor{numclr}{\sffamily #1}}}

@@

@@

Data scientists often use general-purpose programming languages such as Python, but the kind of 
code they write and the way they interact with the system is very different from how software 
engineers work.

This paper focuses on simple data wrangling and data exploration as done, for example, by 
journalists analysing government datasets. This section illustrates how such data analyses 
look and provides a justification for design decisions that we make when designing the data 
exploration calculus.

@@

Data analysts increasingly use notebook systems such as Jupyter or R Markdown, which 
make it possible to combine text, equations and code with results of running the code such as tables 
or visualizations. Notebooks blur the distinction between development and execution. 
Data analysts write small snippets of code, run them to see results immediately and then revise 
them interactively based on the results. We aim to build foundations for such seamless integration 
of development and execution in a notebook environment.

Jupyter notebooks are used by a variety of users ranging from scientists who implement complex
models to journalists who load data, perform simple aggregations and create visualizations. Our
initial focus is on simple use cases, such as those of journalists, because we believe that this 
is where better tooling can provide the most value.

As an example, consider the data anlysis done by the Financial Times for an article on plastic 
waste. The analysis joins datasets from Eurostat, UN Comtrade and more, aggregates the data and 
builds a visualization comparing the plastic waste flows in 2017 and 2018.
Figure~\ref{fig:ft-uncomtrade} shows an extract from one notebook from the data analysis. 
The code of the analysis has a number of remarkable properties:
%
\begin{itemize}
\item There is no abstraction. The full analysis uses lambda functions as arguments to library calls,
  but it does not define any top-level functions. Parameterization is done by defining a list of 
  inputs and then having a for loop, or by defining a top-level variable and setting its value.
  In our example, \ident{material} is set to \str{plastics} and a comment lists other options. 
  This lets the analyst easily see and check the results of intermediate steps of the analysis.

@@

\begin{figure}
\includegraphics[scale=0.6]{notebook.png}
\caption{An excerpt from a Jupyter notebook by Financial Times reporters, which loads trade
  data from the UN trade database and adds ISO country codes.}
\label{fig:ft-uncomtrade}
\end{figure}

@@

\begin{figure}
\includegraphics[scale=0.28]{gamma1.png}
\caption{\small The Gamma script to aggregate Olympic medals. We select Rio 2016 games and
count the number of distinct events per athlete. We then type `.' to choose 
further aggregation operations.}
\label{fig:gamma-dot}
\vspace{-1em}
\end{figure}

@@

\begin{figure}
\includegraphics[scale=0.28 ]{gamma2.png}
\caption{\small Live preview in The Gamma programming environment. The table is produced as
  the data anlyst types and updates on-the-fly to show result at the current cursor position.}
\label{fig:gamma-preview}
\end{figure}

@@

@@

\paragraph{Real-world use cases.}
Our work has been motivated by the needs outlined in the previous section. In particular, we 
have been focusing on a langauge that is simple, yet sufficient for completing basic data 
exploration tasks that a journalist might want to do. We have used our live coding environment
to perform a number of simple data analyses looking at topics such as public government spending, 
crises in financial markets and activities of a research institute. Qualitative observations about
our experience are reported in Section~X. Our tools can be used in any modern web browser and are 
available as an open-source package.

\paragraph{Text editor integration.}
Our approach is based on an ordinary text editor that allows unrestricted modification of the
code. Our parser supports error-recovery and we attempt to give previews whenever the code can be
correctly parsed and type-checked up to the current location. Notably, we do not, in general, know
how the user modified the code. This is in contrast with approaches based on structured editing
where the editor tracks what changes to code have been made. Using an ordinary text editor makes
providing live previews harder, but we believe that many users still favor the flexibility of 
plain text editors.

\paragraph{Data stay, code changes.}
Our focus has been on the scenario when input data is a static file that does not change, but the
data analyst modifies the source code. This is in contrast with work on incremental computation,
which focuses on scenarios where code stays the same, but input data change.

That said, many of the methods that we used for evaluating live previews such as maintaining 
the dependency graph with nodes for individual syntactic elements are closely related to methods
used in incremental computation research. We do not attempt to combine the two, but note that
this is an interesting future direction.

\newpage
\begin{figure}
\raggedright
{\small\sffamily Programs, commands, expressions, arguments and values}
\begin{equation*}
\hspace{-2em}
\begin{array}{lcl}
p &=& c_1; \ldots; c_n\\
c &=& \kvd{let}~x = e \lsep e\\
\end{array}
\quad
\begin{array}{lcl}
e &=& o \lsep x \\
  &|& e.m(a, \ldots, a)
\end{array}
\quad
\begin{array}{lcl}
a &=& e \lsep \lambda x\rightarrow e\\
v &=& o \lsep \lambda x\rightarrow e\\
\end{array}
\end{equation*}

\vspace{0.5em}
{\small\sffamily Evaluation contexts of expressions}
\begin{equation*}
\hspace{-2em}  
\begin{array}{lcl}
C_e[-] &=& -.m(a_1, \ldots, a_n) ~|~ -.m(v_1, \ldots, v_{m}, -, a_1,\ldots , a_n) ~|~ -\\
C_c[-] &=& \kvd{let}~x = C_e[-] ~|~ C_e[-]\\
C_p[-] &=& c_1;\; \ldots;\; c_{k-1};\; C_c[-];\; c_{k+1};\; \ldots;\; c_n\\
\end{array}
\end{equation*}

\vspace{0.5em}
{\small\sffamily Let elimination and member reduction}
\begin{equation*}
\hspace{-2em}  
\begin{array}{ll}
\hspace{-0.5em}\begin{array}{l}  
c_1;\; \ldots;\; c_{k-1};\; \kvd{let}~x=o;\; c_{k+1};\; \ldots; c_n \rightsquigarrow\\
\qquad  c_1;\; \ldots;\; c_{k-1};\; c_{k+1}[x\leftarrow o];\; \ldots; c_n[x\leftarrow o]\end{array} &
\quad\text{\small\sffamily(let)}\\[1.2em]
{o.m(v_1, \ldots, v_n) \rightsquigarrow_\epsilon o'}\implies
  {C_p[o.m(v_1, \ldots, v_n)] \rightsquigarrow C_p[o']} & 
\quad\text{\small\sffamily (external)}    
\end{array}  
\end{equation*}

@@

\subsection{Language syntax}
The calculus combines object-oriented features such as member access with functional features
such as lambda functions. The syntax is defined in Figure~\ref{fig:dec-calculus}. We write $x$ for 
variables and $o$ for object values. A program $p$ in the data exploration calculus consists of a 
sequence of commands $c$. A command can be either let binding or an expression. Lambda functions 
can only appear directly as arguments in method calls. An argument $a$ can thus be a lambda 
function or an expression.

\subsection{Operational semantics}
The data exploration calculus is a call-by-value language. We model the evaluation as a small-step
reduction $\rightsquigarrow$. Fully evaluating a program results in an irreducible sequence of 
objects $o_1;\; \ldots;\; o_n$ which can be displayed as results of the data analysis. The 
operational semantics is parameterized by a relation $\rightsquigarrow_\epsilon$ that models
the functionality of the external libraries used with the calculus and defines reduction rules
for member accesses. The relation has the following form:
%
\begin{equation*}
o.m(v_1, \ldots, v_n) \rightsquigarrow_\epsilon o'
\end{equation*}
%
foo

bar

@@

\subsection{Equational theory}
In The Gamma language, computations are expressed using
member access, written as $e.m(e_1,\ldots, e_n)$. In this paper, we do not define how member
access evaluates. This has been done elsewhere \cite{gamma}, but more importantly, the evaluation of
previews does not rely on the exact specifics of the evaluation, provided that the language
satisfies certain basic conditions. The following definitions provides the necessary structure 
for discussing correctness of previews.

@@

\begin{figure*}
\begin{equation*}
\begin{array}{l}
\xymatrix{
& \bnd{val}(\num{10}) &\qquad\quad& \bnd{val}(\num{15})\\
\bnd{var}(\ident{data}) & 
  \bnd{mem}(\ident{skip},s_0)\ar[l]_{\blbl{arg}(0)} \ar[u]^{\blbl{arg}(1)} && 
  \bnd{mem}(\ident{take},s_1)\ar[ll]_{\blbl{arg}(0)} \ar[u]_{\blbl{arg}(1)}
}\\[5em]
\xymatrix{
&& \bnd{val}(\num{10})\\
\bnd{var}(\ident{data}) & 
  \bnd{mem}(\ident{skip}, s_0)\ar[l]_{\blbl{arg}(0)} \ar@/^/[ru]^{\blbl{arg}(1)} && 
  \bnd{mem}(\ident{take}, s_2)\ar[ll]_{\blbl{arg}(0)} \ar@/_/[lu]_{\blbl{arg}(1)}
}
\end{array}\qquad
\begin{array}{l}
\textnormal{a.) The first graph is constructed from}\\
\textnormal{the following initial expression:}\\[0.5em]
\quad\kvd{let}~x = \num{15}~\kvd{in}\\
\quad\ident{data.skip}(\num{10}).\ident{take}(x)\\
\\
\textnormal{b.) The second diagram shows the updated graph}\\
\textnormal{after the programmer changes $x$ to $10$:}\\[0.5em]
\quad\kvd{let}~x = \num{10}~\kvd{in}\\
\quad\ident{data.skip}(\num{10}).\ident{take}(x)
\end{array}
\end{equation*}
\caption{Dependency graphs formed by two steps of the live programming process. }
\label{fig:dep-graph}
\end{figure*}

@@

In this section, we present a formalisation of a live coding infrastructure for a small,
expression-based programming language that supports \kvd{let} binding, member invocations
and $\lambda$ abstractions. This is the necessary minimum for data exploration as described
in the previous section.

It excludes constructs such as a mechanism for defining new objects as we assume that those
are imported from the context through a mechanism such as type providers.
%
\begin{equation*}
\begin{array}{lcl}
e &=& \kvd{let}~x = e~\kvd{in}~e ~|~ \lambda x\rightarrow e ~|~ e.m(e, \ldots, e) ~|~ x ~|~ n
\end{array}
\end{equation*}
%
Here, $m$ ranges over member names, $x$ over variables and $n$ over primitive values such as 
numbers. Function values can be passed as arguments to methods (provided by a type provider), but 
for the purpose of this paper, we do not need to be able to invoke them directly.

\paragraph{The problem with functions.}
In the context of live programming, \kvd{let} binding and member access are unproblematic.
We can evaluate them and provide live preview for both of them, including all their sub-expressions.
Function values are more problematic, because their sub-expressions cannot be evaluated. For example:
%
\begin{equation*}
\begin{array}{l}
\kvd{let}~\ident{page} = \lambda x \rightarrow \ident{movies}.\ident{skip}(x*\num{10}).\ident{take}(\num{10})
\end{array}
\end{equation*}
%
We can provide live preview for the \ident{movies} sub-expression, but not for 
$\ident{movies}.\ident{skip}(x*\num{10})$ because we cannot obtain the value of $x$ without
running the rest of the program and analysing how the function is called later.

@@

The key idea behind our implementation is to maintain a dependency graph \cite{dependencies} with 
nodes representing individual operations of the computation that can be partially evaluated
to obtain a preview. Each time the program text is modified, we parse it afresh (using an
error-recovering parser) and bind the abstract syntax tree to the dependency graph.

We remember the previously created nodes of the graph. When binding a new expression to
the graph, we reuse previously created nodes that have the same dependencies. For
expressions that have a new structure, we create new nodes (using a fresh symbol to 
identify them).

The nodes of the graph serve as unique keys into a lookup table with previously
evaluated operations of the computation. When a preview is requested, we use the node
bound to the expression to find a preview, or evaluate it by first forcing the evaluation 
of all parents in the dependency graph.

@@

\begin{figure*}[t]
\begin{equation*}
\begin{array}{ll}
\ident{bind}_{\Gamma, \Delta}(e_0.m(e_1, \ldots, e_n)) = &(1)~~\\ 
\qquad \bndclr{v}, (\{\bndclr{v}\} \cup V_0 \cup \ldots \cup V_n, E \cup E_0 \cup \ldots \cup E_n)\\
\quad \textnormal{when}~\bndclr{v_i}, (V_i, E_i) = \ident{bind}_{\Gamma, \Delta}(e_i)\\
\quad \textnormal{and}~\bndclr{v} = \Delta(\bknd{mem}(m),[(\bndclr{v_0}, \blbl{arg}(0)), \ldots, (\bndclr{v_n}, \blbl{arg}(n))])\\
\quad \textnormal{let}~E = \{ (\bndclr{v}, \bndclr{v_0}, \blbl{arg}(0)), \ldots, (\bndclr{v}, \bndclr{v_n}, \blbl{arg}(n))\}
\\[1em]
\ident{bind}_{\Gamma, \Delta}(e_0.m(e_1, \ldots, e_n)) = &(2)\\ 
\qquad \bndclr{v}, (\{\bndclr{v}\} \cup V_0 \cup \ldots \cup V_n, E \cup E_0 \cup \ldots \cup E_n)\\
\quad \textnormal{when}~\bndclr{v_i}, (V_i, E_i) = \ident{bind}_{\Gamma, \Delta}(e_i)\\
\quad \textnormal{and}~(\bknd{mem}(m),[(\bndclr{v_0}, \blbl{arg}(0)), \ldots, (\bndclr{v_n}, \blbl{arg}(n))]) \notin \ident{dom}(\Delta)\hspace{-200em}.\\
\quad \textnormal{let}~\bndclr{v} = \bnd{mem}(m, s), s~\textnormal{fresh}\\
\quad \textnormal{let}~E = \{ (\bndclr{v}, \bndclr{v_0}, \blbl{arg}(0)), \ldots, (\bndclr{v},\bndclr{v_n}, \blbl{arg}(n)) \}
\\[1em]
\ident{bind}_{\Gamma, \Delta}(\kvd{let}~x=e_1~\kvd{in}~e_2) = \bndclr{v}, (\{\bndclr{v}\} \cup V \cup V_1, E \cup E_1)&(3)\\
\quad \textnormal{let}~\bndclr{v_1}, (V_1,E_1) = \ident{bind}_{\Gamma,\Delta}(e_1)\\
\quad \textnormal{let}~\Gamma_1 = \Gamma \cup \{(x,\bndclr{v_1})\} \\
\quad \textnormal{let}~\bndclr{v}, (V, E) = \ident{bind}_{\Gamma_1, \Delta}(e_2)
\end{array}
\begin{array}{ll}
\ident{bind}_{\Gamma, \Delta}(n) = \bnd{val}(n), (\{ \bnd{val}(n) \}, \emptyset) }&(4)
\\[1em]
\ident{bind}_{\Gamma, \Delta}(x) = \bndclr{v}, (\{ \bndclr{v} \}, \emptyset)\quad \textnormal{when}~\bndclr{v} = \Gamma(x)&(5)
\\[1.25em]
\ident{bind}_{\Gamma, \Delta}(\lambda x \rightarrow e) = \bndclr{v}, (\{\bndclr{v}\} \cup V, \{e\} \cup E)&(6)\\
\quad \textnormal{when}~\Gamma_1 = \Gamma \cup \{ x, \bnd{var}(x) \}\\
\quad \textnormal{and}~\bndclr{v_0}, (V, E) = \ident{bind}_{\Gamma_1, \Delta}(e)\\
\quad \textnormal{and}~\bndclr{v} = \Delta(\bknd{fun}(x),[(\bndclr{v_0}, \blbl{body})])\\
\quad \textnormal{let}~e = (\bndclr{v}, \bndclr{v_0}, \blbl{body}) 
\\[1.25em]
\ident{bind}_{\Gamma, \Delta}(\lambda x \rightarrow e) = \bndclr{v}, (\{\bndclr{v}\} \cup V, \{e\} \cup E)&(7)\\
\quad \textnormal{when}~\Gamma_1 = \Gamma \cup \{ x, \bnd{var}(x) \}\\
\quad \textnormal{and}~\bndclr{v_0}, (V, E) = \ident{bind}_{\Gamma_1, \Delta}(e)\\
\quad \textnormal{and}~(\bknd{fun}(x),[(\bndclr{v_0}, \blbl{body})])\notin \ident{dom}(\Delta)\\
\quad \textnormal{let}~\bndclr{v} = \bnd{fun}(x, s), s~\textnormal{fresh}\\
\quad \textnormal{let}~e = (\bndclr{v},\bndclr{v_0},\blbl{body}) 
\\[0.75em]
\end{array}
\end{equation*}
\caption{Rules of the binding process, which constructs a dependency graph for an expression.}
\label{fig:binding-rules}
\end{figure*}

% --------------------------------------------------------------------------------------------------
\paragraph{Elements of the graph.} The nodes of the graph represent individual operations
to be computed. In our design, the nodes themselves are used as keys, so we attach a unique 
\emph{symbol} to some of the nodes. That way, we can create two unique nodes representing, 
for example, access to a member named \ident{take} which differ in their dependencies.

>>>>

\usepackage{booktabs}
\usepackage{subcaption}
\usepackage{lineno,hyperref,xcolor}
\usepackage{flushend}
\usepackage{stmaryrd}
\usepackage{amssymb}
\usepackage{xypic}
\usepackage{semantic}
\usepackage{booktabs}
\usepackage{enumitem}
\usepackage{enumerate}
\usepackage{amsmath}

@@

\definecolor{cmtclr}{rgb}{0.0,0.6,0.0}
\definecolor{kvdclr}{rgb}{0.0,0.0,0.6}
\definecolor{idclr}{rgb}{0.3,0.3,0.3}
\definecolor{numclr}{rgb}{0.0,0.4,0.0}
\definecolor{strclr}{rgb}{0.4,0.4,0.0}
\definecolor{rstrclr}{rgb}{0.5,0.1,0.0}
\definecolor{prepclr}{rgb}{0.6,0.0,0.2}
\newcommand{\vect}[1]{\langl #1 \rangl}
\newcommand{\langl}{\begin{picture}(4.5,7)
\put(1.1,2.5){\rotatebox{60}{\line(1,0){5.5}}}
\put(1.1,2.5){\rotatebox{300}{\line(1,0){5.5}}}
\end{picture}}
\newcommand{\rangl}{\begin{picture}(4.5,7)
\put(.9,2.5){\rotatebox{120}{\line(1,0){5.5}}}
\put(.9,2.5){\rotatebox{240}{\line(1,0){5.5}}}
\end{picture}}
\newcommand{\ball}[1]{\FPeval{\result}{clip(201+#1)}\textnormal{\ding{\result}}}
\newcommand{\lsep}{\;\;|\;\;}
\newcommand{\num}[1]{\textcolor{numclr}{#1}}
\newcommand{\str}[1]{\textnormal{\textcolor{strclr}{\sffamily "#1"}}}
\newcommand{\rstr}[1]{\textnormal{\textcolor{rstrclr}{\sffamily "#1"}}}
\newcommand{\ident}[1]{\textnormal{\textcolor{idclr}{\sffamily #1}}}
\newcommand{\qident}[1]{\textnormal{\sffamily \textquotesingle #1\textquotesingle}}
\newcommand{\dom}{\ident{dom}}
\newcommand{\kvd}[1]{\textnormal{\textcolor{kvdclr}{\sffamily #1}}}

\newcommand{\bndclr}[1]{\textcolor{kvdclr}{#1}}
\newcommand{\bkndclr}[1]{\textcolor{prepclr}{#1}}
\newcommand{\blblclr}[1]{\textcolor{numclr}{#1}}
\newcommand{\bnd}[1]{\textnormal{\textcolor{kvdclr}{\sffamily #1}}}
\newcommand{\bknd}[1]{\textnormal{\textcolor{prepclr}{\sffamily #1}}}
\newcommand{\blbl}[1]{\textnormal{\textcolor{numclr}{\sffamily #1}}}
\newcommand{\narrow}[1]{\hspace{-0.6em}#1\hspace{-0.6em}}

@@

\begin{figure}[b]
\includegraphics[scale=0.5]{notebook.png}
\caption{An excerpt from a Jupyter notebook by Financial Times reporters, which loads trade
  data from the UN trade database and adds ISO country codes.}
\label{fig:ft-uncomtrade}
\end{figure}

@@

Data scientists often use general-purpose programming languages such as Python, but the kind of
code they write and the way they interact with the system is very different from how software
engineers work. This paper focuses on simple data wrangling and data exploration as done, for
example, by journalists analysing government datasets. This section illustrates how such data
analyses look and provides a justification for the design of our data exploration calculus.

@@

Data analysts increasingly use notebook systems such as Jupyter or R Markdown, which
make it possible to combine text, equations and code with results of running the code such as tables
or visualizations. Notebooks blur the distinction between development and execution.
Data analysts write small snippets of code, run them to see results immediately and then revise
them.

% interactively based on the results. We aim to build foundations for such seamless integration
% of development and execution in a notebook environment.

Jupyter notebooks are used by a variety of users ranging from scientists who implement complex
models to journalists who load data, perform simple aggregations and create visualizations. Our
initial focus is on simple use cases, such as those of journalists. Tooling that makes notebook
systems more interactive and spreadsheet-like in such cases is of crucial importance as it enables
more people to produce reproducible and transparent data analyses.

As an example, consider the anlysis done by the Financial Times for an article on plastic
waste. The analysis joins datasets from Eurostat, UN Comtrade and more, aggregates the data and
builds a visualization comparing plastic waste flows in 2017 and 2018.
Figure~\ref{fig:ft-uncomtrade} shows an extract from one notebook from the data analysis.
The code of the analysis has a number of remarkable properties:
%
\begin{itemize}
\item There is no abstraction. The full analysis uses lambda functions as arguments to library calls,
  but it does not define any top-level functions. Parameterization is done by defining a list of
  inputs and then having a for loop, or by defining a top-level variable and setting its value.
  In our example, \ident{material} is set to \str{plastics} and a comment lists other options.
  This lets the analyst easily see and check the results of intermediate steps of the analysis.

@@

@@

\begin{figure}[b]
  \centering
  \subcaptionbox{
    \raggedright\footnotesize The Gamma script to aggregate Olympic medals. We select Rio 2016 games and
    count the number of distinct events per athlete. We then type `.' to choose
    further aggregation operations.
    \label{fig:gamma-dot}
  }{\includegraphics[scale=0.26]{gamma1.png}}
  \subcaptionbox{
    \raggedright\footnotesize Live preview in The Gamma programming environment. The table is produced as
      the data anlyst types and updates on-the-fly to show result at the current cursor position.
    \label{fig:gamma-preview}
  }{\includegraphics[scale=0.26]{gamma2.png}}
  \caption{\small TBD}
  \label{first-figure-with-subfigures}
\end{figure}

@@

@@

% TODO: What is live coding? (Show preview while typing!)

@@

\begin{itemize}
\item\textbf{Real-world use cases.}
Our work has been motivated by the needs outlined in the previous section. In particular, we
have been focusing on a langauge that is simple, yet sufficient for completing basic data
exploration tasks that a journalist might want to do. We have used our live coding environment
to perform a number of simple data analyses looking at topics such as public government spending,
crises in financial markets and activities of a research institute. Qualitative observations about
our experience are reported in Section~X. Our tools can be used in any modern web browser and are
available as an open-source package.

\item\textbf{Text editor integration.}
Our approach is based on an ordinary text editor that allows unrestricted modification of the
code. Our parser supports error-recovery and we attempt to give previews whenever the code can be
correctly parsed and type-checked up to the current location. Notably, we do not, in general, know
how the user modified the code. This is in contrast with approaches based on structured editing
where the editor tracks what changes to code have been made. Using an ordinary text editor makes
providing live previews harder, but we believe that many users still favor the flexibility of
plain text editors.

\item\textbf{Data stay, code changes.}
Our focus has been on the scenario when input data is a static file that does not change, but the
data analyst modifies the source code. This is in contrast with work on incremental computation,
which focuses on scenarios where code stays the same, but input data change.

That said, many of the methods that we used for evaluating live previews such as maintaining
the dependency graph with nodes for individual syntactic elements are closely related to methods
used in incremental computation research. We do not attempt to combine the two, but note that
this is an interesting future direction.
\end{itemize}

\newpage
\begin{figure}
\raggedright
{\small\sffamily Programs, commands, expressions, arguments and values}
\begin{equation*}
\hspace{-2em}
\begin{array}{lcl}
p &\narrow{=}& c_1; \ldots; c_n\\
c &\narrow{=}& \kvd{let}~x = e \lsep e\\
\end{array}
\qquad
\begin{array}{lcl}
e &\narrow{=}& o \lsep x \\
  &\narrow{|}& e.m(a, \ldots, a)
\end{array}
\qquad
\begin{array}{lcl}
a &\narrow{=}& e \lsep \lambda x\rightarrow e\\
v &\narrow{=}& o \lsep \lambda x\rightarrow e\\
\end{array}
\end{equation*}

\vspace{0.5em}
{\small\sffamily Evaluation contexts of expressions}
\begin{equation*}
\hspace{-2em}
\begin{array}{lcl}
C_e[-] &=& -.m(a_1, \ldots, a_n) ~\lsep~ -.m(v_1, \ldots, v_{m}, -, a_1,\ldots , a_n) ~\lsep~ -\\
C_c[-] &=& \kvd{let}~x = C_e[-] \hspace{1.25em}\lsep~ C_e[-]\\
C_p[-] &=& o_1;\; \ldots;\; o_{k};\; C_c[-];\; c_{1};\; \ldots;\; c_n\\
\end{array}
\end{equation*}

\vspace{0.5em}
{\small\sffamily Let elimination and member reduction}
\begin{equation*}
\hspace{-2em}
\begin{array}{ll}
\hspace{-0.5em}\begin{array}{l}
o_1;\; \ldots;\; o_{k};\; \kvd{let}~x=o;\; c_{1};\; \ldots; c_n \rightsquigarrow\\
\qquad  o_1;\; \ldots;\; o_{k};\; c_{1}[x\leftarrow o];\; \ldots; c_n[x\leftarrow o]\end{array} &
\quad\text{\small\sffamily(let)}\\[1.2em]
{o.m(v_1, \ldots, v_n) \rightsquigarrow_\epsilon o'}\implies
  {C_p[o.m(v_1, \ldots, v_n)] \rightsquigarrow C_p[o']} &
\quad\text{\small\sffamily (external)}
\end{array}
\end{equation*}

@@

\subsection{Language syntax}
The calculus combines object-oriented features such as member access with functional features
such as lambda functions. The syntax is defined in Figure~\ref{fig:dec-calculus}. Object values
$o$ are defined by external libraries that are used in conjunction with the core calculus.

A program $p$ in the data exploration calculus consists of a sequence of commands $c$. A command
can be either let binding or an expression. Let bindings define variables $x$ that can be used in
subsequent commands. The calculus supports lambda functions, but they
can only appear as arguments in method calls. An argument $a$ can thus be a lambda
function or an expression.\footnote{Similar but weaker restrictions on the use of lambda functions
exist in other languages. To guide type inference, lambda functions in C\# can appear as either
method arguments or assigned to an explicitly typed variable.}

\subsection{Operational semantics}
The data exploration calculus is a call-by-value language. We model the evaluation as a small-step
reduction $\rightsquigarrow$. Fully evaluating a program results in an irreducible sequence of
objects $o_1;\; \ldots;\; o_n$ which can be displayed as intermediate results of the data analysis. The
operational semantics is parameterized by a relation $\rightsquigarrow_\epsilon$ that models
the functionality of the external libraries used with the calculus and defines reduction rules
for member accesses. The relation has the following form:
%
\begin{equation*}
o_1.m(v_1, \ldots, v_n) \rightsquigarrow_\epsilon o_2
\end{equation*}
%
Figure~\ref{fig:dec-calculus} defines the reduction rules in terms of $\rightsquigarrow_\epsilon$
and evaluation contexts; $C_e$ specifies left-to-right evaluation of arguments of a method call,
$C_c$ specifies evluation of a command and $C_p$ defines top-to-bottom evaluation of a program.
The rule {\small\sffamily (external)} applies reductions provided by an external library in a
call-by-value order and {\small\sffamily (let)} substitutes a value of an evaluated variable in
all subsequent commands.

\subsection{Properties}

TBD: Some adaptation of the stuff below, as needed for the proofs.

@@

% TODO: Adapt as needed
%
% \subsection{Equational theory}
% In The Gamma language, computations are expressed using
% member access, written as $e.m(e_1,\ldots, e_n)$. In this paper, we do not define how member
% access evaluates. This has been done elsewhere \cite{gamma}, but more importantly, the evaluation of
% previews does not rely on the exact specifics of the evaluation, provided that the language
% satisfies certain basic conditions. The following definitions provides the necessary structure
% for discussing correctness of previews.
%
% Partial evaluation may reduce an expression under $\lambda$-abstraction. We do not require that
% the reduction of the host language does this. Instead, we define an extended reduction relation
% and use that in the proofs. The host language only needs to compose well with such extended reduction
% as captured by the \emph{compositionality} property below. We also require that the language allows
% elimination of let bindings.
%
% \begin{definition}[Host language]
% \label{def:host}
% Given a relation on expressions $e_1 \rightsquigarrow e_2$ that models small-step evaluation, we define:
%
% \begin{itemize}
% \item[--]
%  A \emph{preview evaluation context} (also referred to as \emph{context}):
%
% \vspace{-1.25em}
% \begin{equation*}
% \begin{array}{lcl}
% C[-] &\hspace{-1em}=\hspace{-1em}& \kvd{let}~x = -~\kvd{in}~e ~|~ \kvd{let}~x = e~\kvd{in}~- ~|~ \lambda x\rightarrow - \\
%      &&  e_0.m(e_1, \ldots, e_{k-1}, -, e_{k+1}, \ldots e_n)
% \end{array}
% \end{equation*}
%
% \item[--]
% An extended reduction relation $\rightsquigarrow_\beta$ such that, for any context $C$,
% $C[e_1] \rightsquigarrow_\beta C[e_2]$ whenever $e_1 \rightsquigarrow e_2$.
%
% \item[--]
% Let elimination $\rightsquigarrow_\ident{let}$ such that, using  capture-avoiding substitution,
%   $C[\kvd{let}~x=e_1~\kvd{in}~e_2] \rightsquigarrow_\ident{let} C[e_2[x\leftarrow e_1]]$
% \end{itemize}
%
% \noindent
% We say that $\rightsquigarrow$ is a suitable \emph{host language reduction} if:
% \begin{itemize}
% \item[--] It satisfies the \emph{compositionality} property, that is if $e \rightsquigarrow e'$ and
% $C[e]\rightsquigarrow_\beta e''$ then also $C[e']\rightsquigarrow_\beta e''$.
%
% \item[--] Let elimination does not affect the result, i.e.~if $e \rightsquigarrow_{\ident{let}} e'$
% and $e'\rightsquigarrow_\beta e''$ then also $e\rightsquigarrow_\beta e''$
% \end{itemize}
% \end{definition}
%
% \noindent
% The host language in The Gamma is a simple call-by-value functional language without side-effects,
% and so it satisfies both compositionality and allows let bindings to be eliminated, although the
% latter affects the performance. The mechanism for preview evaluation presented here would also
% work for call-by-name languages, but it would suffer from the expected difficulties in the
% presence of side-effects or non-determinism.
%
% \newpage

@@

@@

A naive way of providing live previews during code editing would be to re-evaluate the code
after each change. This would be wasteful -- for example, when writing
code, most changes are additive and the preview can be updated by evaluating just the newly added
code. In this section, we develop foundations for an efficient way of displaying live previews
for the data exploration calculus.

@@

The key idea behind our method is to maintain a dependency graph \cite{dependencies} with
nodes representing individual operations of the computation that can be (partially) evaluated
to obtain a preview. Each time the program text is modified, we parse it afresh (using an
error-recovering parser) and bind the abstract syntax tree to the dependency graph.
When binding a new expression to the graph, we reuse previously created nodes as long as
they have the same dependencies. For expressions that have a new structure, we create new nodes.

The nodes of the graph serve as unique keys into a lookup table with previously
evaluated parts of the program. When a preview is requested for an expression, we use the graph
node bound to the expression to find a preview. If a preview has not been evaluated, we force
the evaluation of all dependencies in the graph and then evaluate the operation represented by
the current node.

@@

\begin{figure}[b]
\centering
\subcaptionbox{
  \raggedright\small
  Graph constructed from the following initial expression:
  $\kvd{let}~x = \num{15}~\kvd{in}~\ident{data.skip}(\num{10}).\ident{take}(x)$
  \label{fig:dep-graph-a}
}{
  $\xymatrix{
  \bnd{val}(\num{10}) & \bnd{val}(\num{15})\\
  \bnd{mem}(\ident{skip},s_0)\ar[d]_{\blbl{arg}(0)} \ar[u]^{\blbl{arg}(1)} &
  \bnd{mem}(\ident{take},s_1)\ar[l]_{\blbl{arg}(0)} \ar[u]_{\blbl{arg}(1)}\\
  \bnd{var}(\ident{data})
  }\qquad$
}
\hspace{1em}
\subcaptionbox{
  \raggedright\small
  Updated graph after changing $x$ to $10$:
  $\kvd{let}~x = \num{10}~\kvd{in}~\ident{data.skip}(\num{10}).\ident{take}(x)$
  \label{fig:dep-graph-b}
}{
  $\xymatrix{
  \bnd{val}(\num{10})&\\
  \bnd{mem}(\ident{skip}, s_0)\ar[d]_{\blbl{arg}(0)} \ar[u]^{\blbl{arg}(1)} &
  \bnd{mem}(\ident{take}, s_2)\ar[l]_{\blbl{arg}(0)} \ar@/_/[lu]_{\blbl{arg}(1)}\\
  \bnd{var}(\ident{data})
}\qquad$
}
\caption{Dependency graphs formed by two steps of the live programming process. }
\label{fig:dep-graph}
\end{figure}
